---
title: "CA4"
author: " Laylaa Jorge Abrahams, Roche Witbooi, Chris Eason"
format: "pdf"
---

# Question 1

Suppose X ∼ Np(µ,Σ). Show that the maximum likelihood estimator of Σ is biased, and give the bias:

1.  The following shows that the MLE of $\Sigma$ is biased because of the added $\frac{n-1}{n}$ term

$$
E[\hat{\Sigma}_{MLE}] = E[\frac{n-1}{n}S] = \frac{n-1}{n}E[S] = \frac{n-1}{n}\Sigma
$$

2.  The bias of the MLE of $\Sigma$

$$
Bias(\hat{\Sigma}_{MLE}) = E[\hat{\Sigma}_{MLE}] - \Sigma
$$

$$
Bias(\hat{\Sigma}_{MLE}) = \frac{n-1}{n}\Sigma - \Sigma
$$

$$
Bias(\hat{\Sigma}_{MLE}) = -\frac{1}{n}\Sigma
$$

\newpage

# Question 2

Consider again the Egyptian skull data from CA1, given in CA1.csv . Examine the variables in period 2 for marginal and multivariate normality by creating the necessary QQ-plot(s) and chi-square plot(s). Apply any statistical test to the univariate hypotheses and report a measure of the p-value. For the multivariate test, interpret the observed squared generalized distances.

```{r, echo=FALSE, hide= TRUE}
dat <- read.csv("CA1.csv")

library(dplyr)
period_2 <- dat |> filter(TimePeriod ==2)
period_2 <- scale(period_2[,1:4])
```

```{r, echo=FALSE, hide= TRUE}
library(ggplot2)
# chi-squared plots
d2 <- mahalanobis(period_2, colMeans(period_2), cov= cov(period_2))
d2 <- sort(d2)
qcp <-  qchisq((1:nrow(period_2) - 0.5)/nrow(period_2), ncol(period_2))

df1 <- data.frame(x=qcp, y=d2)
ggplot()+
  geom_point(data = df1, aes(x=x, y=y), color = "black")+
  geom_abline(intercept = 0, slope = 1, color="red") +
  labs(
    title = "Chi-Squared plot",
    subtitle = "Theoretical quantiles from a Chi-Squared with P = 4 degrees
of freedom (QCP) and Mahalanobis Distance", 
    x = "QCP",
    y = "Mahalanobis Distance"
  ) +
  scale_x_continuous(
    breaks = seq(0, 14, by = 2), 
    limits = c(0, 14)  
  ) +
  scale_y_continuous(
    breaks = seq(0, 14, by = 2),  
    limits = c(0, 14)  
  ) +
  coord_fixed(ratio = 1) +  
  theme_minimal()

critical <- qchisq(0.5, df = 4)
print(paste("Proportion of points where D^2 < critical value", round(mean(d2 < critical)*100),2))

```

```{r, echo=FALSE, hide= TRUE}
library(knitr)
library(ggplot2)
library(patchwork)

# qqplot
p_val <- matrix(nrow=4,ncol = 1)
plots <- list()

for(i in 1:ncol(period_2)){
  result <- qqnorm(period_2[, i], main=colnames(period_2)[i], plot.it = FALSE)
  df <- data.frame(x = result$x, y = result$y)
  plots[[i]] <- ggplot()+
    geom_point(data = df, aes(x=x,y=y), color = "black")+
    geom_abline(slope = 1, intercept = 0, color = "red") +
    ggtitle(colnames(period_2)[i]) +
    coord_fixed(ratio = 1) +
    scale_x_continuous(
      breaks = seq(-4, 4, by = 2),  
      limits = c(-4, 4)  
    )+
    scale_y_continuous(
      breaks = seq(-4, 4, by = 2),  
      limits = c(-4, 4)  
    ) + 
    labs(
      y = "Standardised \nObserved\n Quantiles", 
      x = "Theoretical\n Expected\n Quantiles"
    ) +
    theme_minimal() +
    theme(
      axis.title.x = element_text(size = 5),
      axis.title.y = element_text(size = 5), 
      plot.title = element_text(size = 8), 
      ) 
  
  p_val[i] <- shapiro.test(period_2[,i])$p.value
}
wrap_plots(plots, nrow=2) + 
  plot_layout(nrow = 2, ncol = 2, widths = c(2, 2), heights = c(2, 2))
rownames(p_val) <- colnames(period_2)
colnames(p_val) <- "P Value"

kable(p_val, caption = "Shapiro Wilk Test")
```
